---
title: "Using DotMatch from R"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Using DotMatch from R}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Introduction

`dotmatch` is a fast, deterministic tool for assigning short DNA sequences (CRISPR guides, feature barcodes, inline barcodes, etc.) against a known whitelist while explicitly reporting ambiguous cases.

This R package provides thin, convenient wrappers around the Python implementation using `reticulate`. For very large datasets, we strongly recommend calling the native `dotmatch` CLI (installed via Bioconda or from source) and then reading the TSV/JSON outputs into R (or using `zellkonverter` + AnnData for SingleCellExperiment interop).

## Installation

```r
# reticulate + the R package
install.packages("reticulate")
# Then in Python (recommended via reticulate::py_install or a conda env):
# pip install "dotmatch[anndata]"

# The R package itself can be installed from source:
# devtools::install("path/to/dotmatch/R")
```

## Basic Usage

```{r setup}
library(dotmatch)
```

Compute distances:

```{r}
distance("ACGT", "AGGT")
```

Assign a batch (small data):

```{r}
targets <- c("ACGT", "AGGT", "ACGA")
reads   <- c("ACGT", "ACGC", "TTTT")
res <- assign(reads, targets, k = 1)
print(res)
```

Using the Matcher for repeated assignments (efficient for many reads vs fixed library):

```{r}
m <- Matcher(c("ACGT", "AGGT", "ACGA"))
res2 <- m$assign(c("ACGT", "ACGC"))
print(res2)
m$close()
```

Feature assignment from df (common for cell x feature-seq after extraction):

```{r, eval=FALSE}
# df has column 'feature_seq'
# library can be path to tsv or df
# df <- assign_features_df(df, seq_col = "feature_seq", library = "features.tsv", k = 1)
```

For AnnData / feature count matrices, use the Python helpers via reticulate or convert outputs from the CLI.

## Recommended Production Workflow (Industry / Large Screens)

1. Use the native CLI (or `dotmatch assay run` with TOML spec) for speed, determinism, and rich QC artifacts.
2. Read `counts.mageck.tsv` or `assignments.tsv` + `sample_qc.tsv` into R.
3. Use `zellkonverter::AnnData2SingleCellExperiment` if you went through Python anndata helpers, or construct manually with `SingleCellExperiment`.

This gives you the best of both worlds: C-speed + exact semantics + full R/Bioconductor ecosystem.

See the main DotMatch documentation for the full set of CLI commands, schemas, and evidence of correctness.
